A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-22 latex/main.tex
Updates from Overleaf

By automatically detecting and incorporating recombination events into the ARG,
sc2ts provides us with a systematic means of
discovering and assessing evidence around these events.
Table~\ref{tab:pango_x_lineages} summarizes
these results for the 44 Pango X lineages present in the ARG
(XAC and XAD have been omitted, and XM partially represented,
corresponding to 23 omitted samples in total;
see Document~\ref{sec:supplement_x_pango_lineages}).
For 16 Pango X lineages, the oldest node
that has been assigned to the focal Pango lineage is a recombination node, and
that node is the ancestor of all samples
assigned to that (or a descendant) Pango lineage.
These are the Type I events, in which the ARG reflects the
hierarchical structure of Pango designation.
Among these events, concordance with the community-consensus 
parent lineages and breakpoint intervals is excellent,
with similar performance to
the RecombinHunt \cite{Alfonsi2024}
and CovRecomb \cite{Li2024CovRecomb} detection methods
(STAR Methods; Table~\ref{tab:method_concordance}; see also
Document~\ref{sec:jackson_recombs} for analysis of the 
Jackson et al.\ recombinants).
Of the Type I events, 13 are monophyletic for the focal Pango lineage
(i.e., the descendants are entirely from that lineage);
Figure~\ref{fig:subgraphs}A shows XA as an example.

@@

We validated the recombination events in the sc2ts ARG by running 
3SEQ~\cite{Lam2018improved} on the putative recombinant 
sequence trios (STAR Methods), 
and 484 of the 855 (57\%) were classified as recombinant.
Validation status correlates strongly with our QC filter, with 
95\% of the QC-passing recombinants 
and 29\% of the QC-failing recombinants
being classified as recombinant.
In addition to 3SEQ, we ran the CovRecomb~\cite{Li2024CovRecomb} 
and rebar recombination detection methods on the putative recombinant 
sequences, summarized in Table~\ref{tab:recombinant_validation}.
CovRecomb has a substantially lower validation rate at 25\% 
among the QC-passing recombinants. This is expected, as 
CovRecomb and is specifically intended to detect interlineage recombinants
and been previously shown to be
less sensitive than other methods~\cite{Li2024CovRecomb}.
Rebar validation is intermediate, at 57\% among QC-passing recombinants.
(See also Figures~\ref{fig:recombinant_evidence_cov_recomb},
and \ref{fig:recombinant_evidence_rebar}.)
% We don't talk about RIPPLES results yet as we haven't discussed it, a
% and that would be confusing.

@@

Recombination is rare in SARS-CoV-2, and
ultimately judgments about plausibility require expert analysis.
To aid such interpretation and help identify quality control issues
we developed a visualization tool
inspired by SnpIt~\cite{OToole2024} and 
the recombinant view of RIVET~\cite{Smith2023tracking},
showing the recombination-informative sites
between a recombinant and its parent sequences, and
highlighting potentially problematic loci.
Each copying pattern shows the positions
where the allelic state in a recombinant
(middle row) matches that of the left (upper row, colored green) 
or right parent (lower row, colored blue),
or where the recombination event requires a \emph{de novo} mutation
(gold, with mutational change below).
Figure~\ref{fig:recombinant_evidence}D shows some example copying patterns,
illustrating this spectrum of plausibility.
XA is an example of a highly plausible recombinant, shown by the 
large number of supporting loci on the left and right. As we have 
only one \emph{de novo} mutation, we can be sure that the parents are closely
observed [Some details about XA here, how well the parents were sampled etc].
XBB is also strongly supported by many loci on the left and right, 
but the presence of many \emph{de novo} mutations
observed in the copying patterns suggests the existence of
many genetically distinct, unsampled ancestors 
[FIXME: in this case, probably BJ.1, see issue 336],
and some caution should be exercised in interpreting them. 
See Figures~\ref{fig:recombinant_evidence}D and
\ref{fig:pango_x_copying_patterns}
for copying patterns of the Pango X lineages, and
Figure~\ref{fig:copying_pattern_examples} for copying patterns
highlighting specific types of QC problems
(and more details on the visualizations).

@@

In terms of \emph{detecting} recombination events, 
sc2ts and the combination of UShER and RIPPLES have much in common, 
and it is likely that very similar results could be obtained if run
in the same way. That is, if we constructed an UShER tree by 
incrementally adding the same daily batches as used for sc2ts
and examined any newly-added long branches each day
using RIPPLES, we would likely get very similar results to sc2ts
in terms recombinants detected.
For a given matching operation, RIPPLES and the LS HMM 
(under its current parameterisation) 
are deterministically solving the same well-defined problem
(albeit with radically different implementations) and they 
must therefore produce similar answers. 
Where sc2ts and UShER+RIPPLES diverge, however, is in terms 
of how well such signals of recombination can be preserved
over longer timescales. Continuing our thought experiment
in which we run UShER and RIPPLES in daily batches, suppose
that on a given day a branch with 4 mutations is identified 
as being a recombinant. Unless this branch is specifically 
``protected'' as representing a recombinant it may 
be broken up by subsequent parsimony operations, thus 
erasing the signal of recombination. This would appear
to be the mostly likely explanation for the ``missing'' 
X-lineages like XBB (detected by sc2ts at $k=4$ but 
not by UShER+RIPPLES at $p=4$).

@@

% P2. This is a conservative lower-bound on recombination. 
% Discuss other estimates.

The recombination frequency estimated from the \texttt{sc2ts} ARG is
substantially lower than that implied by the combination of UShER+RIPPLES,
largely reflecting differences in model formulation and parameter choice.
RIPPLES, applied post-hoc to UShER trees of 1.6 million genomes up to May 2021,
reported 589 unique recombination events with 43,104 descendant
samples---corresponding to roughly 2.7\% of all genomes being of recombinant
origin.  Using the same time interval, our inference at \(k=4\) identifies only
14 high-confidence recombination events with 172 descendants among 501,235
genomes (0.034\%), a 40- to 80-fold reduction.  This difference is largely
explained by the stricter parsimony penalty adopted here: increasing the
penalty from \(p=3\) (as used in the RIPPLES analysis) 
to \(k=4\) greatly reduces false
positives while retaining nearly all verified recombinant lineages, 
including those missed by RIPPLES at equivalent thresholds.  
We argue that the post-hoc nature of UShER+RIPPLES can erase evidence 
of recombination as
tree rearrangements proceed, thereby requiring the more lenient 
threshold in order to find known recombinats.
These comparisons suggest that our estimate represents a conservative 
lower bound. Further work will be required to determine
the true frequency of recombination in SARS-CoV-2 evolution.

% P3. Clarifies relationships between pango X lineages, simplifying
% the Omicron sublineage.

@@

% P4. Results on global case numbers and frequency of recomb.

[FIXME: This is direct output from ChatGPT.  Review!]
The dated ARG allows us to place recombination within its epidemiological
context, revealing clear temporal structure in when recombinant lineages arise.
The frequency of inferred recombination events follows global case numbers
through the pandemic, with peaks during the Delta and Omicron waves when both
infection prevalence and genetic diversity were high.  Quantitatively, the best
single predictor of weekly recombinant counts is the product of global cases
and lineage heterozygosity (\(nH\)), which explains nearly a quarter of the
variation in event frequency (\(R^{2}_{\mathrm{adj}} = 0.24\)).  This
relationship is consistent with the mechanistic expectation that recombination
requires both frequent co-infections (driven by high case numbers) and
genetically distinct co-infecting viruses (driven by lineage diversity).  In
other words, recombination becomes most detectable when epidemiological and
evolutionary opportunity coincide.  These results imply that the emergence of
novel recombinants is not uniform through time but concentrated into
predictable ``hot periods'' during large, antigenically diverse outbreaks, when
the ecological conditions for viral recombination are most favourable.

% P5. Confirm breakpoints around Spike.

[FIXME: This is directish output from ChatGPT. Review!]
The genomic distribution of breakpoints in the ARG is highly uneven, with a
pronounced enrichment near the $5'$ boundary of the Spike gene and across the
S1–S2 junction.  This pattern closely mirrors that observed in the
UShER+RIPPLES dataset \citep{Turakhia2022} and in earlier 
analyses \citep{Jackson2021} [more cites], but here it is recovered
independently over a much longer sampling period and under a more conservative
inference regime.  Under our more conservative estimate of the overall
prevalence of recombination, we find that this enrichment persists despite a substantial
reduction in the total number of inferred recombination events, indicating that
it reflects a genuine and stable feature of SARS-CoV-2 evolution rather than a
by-product of detection sensitivity.  Across all high-confidence events, the
median breakpoint interval length is $\sim 2.2$kb and declines with parental
divergence, consistent with expectations from sequence-based localization
accuracy.  These results show that even with a conservative lower bound on
recombination frequency, the concentration of breakpoints around Spike remains
a robust signature of coronavirus genome biology—likely reflecting a
combination of mechanistic biases in template switching and selective
constraints on Spike function.

% P6. Sc2ts method is a major advance. We validate with real data
% rather than simulations. Forward looking statement about simulations.

In the broader context of pathogen genomics, \texttt{sc2ts} demonstrates that
it is now possible to reconstruct recombination-aware genealogies at the scale
of entire pandemics.  Although recombination is recognised as a major
evolutionary force in many pathogens—including RNA viruses such as influenza
and HIV, and numerous bacterial species—genealogical analyses have typically
relied on tree-based approximations that ignore or collapse recombination.
This simplification has limited the accuracy of downstream evolutionary and
epidemiological inference, even in well-studied systems.  \texttt{Sc2ts}
overcomes this barrier by directly inferring an ancestral recombination graph
from millions of viral genomes, producing the first coherent reticulate
genealogy for any pathogen dataset of comparable size.  The ability to
integrate recombination into genome-wide genealogies at this scale represents a
major conceptual step forward: it transforms recombination from a confounding
complication into an analyzable component of pathogen evolution.

@@

We close by emphasizing several current limitations and potential directions to improve sc2ts.
First, more stringent filters to exclude false positive recombinants during inference
would obviate post-hoc treatment of these recombinants.
Second, indels carry important phylogenetic signals,
for example, as lineage-defining mutations of Alpha and Delta (REF).
We currently treat multi-base deletions as missing data or multiple independent deletions,
but they are in reality single events.
More sophisticated treatments of multi-base indels in the HMM could help recover additional signals
(and avoid false signals) of recombination.
Third, the current HMM assumes a simplistic mutation model,
which does not account for documented transition-to-transversion rate bias (REF)
and context-dependent variation in mutation rate (REF).
A more realistic mutation model may help improve inference of local genetic relationships in the ARG.
Fourth, an inherent limitation of detecting recombination from sequence data,
which applies to all methods, is that it is impossible to detect events with few (or no) supporting bases,
as these cannot be distinguished from mutation events.
This limitation makes it particularly challenging to detect events near the termini of linear viral genomes.
Fifth, it is challenging to accurately reconstruct global events
when there are regions with poor sampling and jurisdictions with long delays
before sequences become publicly available (REF).
Revisiting the ARG topology periodically
(e.g., rewinding the ARG a month and repeating with samples that became available later)
and better handling of putative recombinants with suspiciously long branches
(e.g., Delta and Omicron BA.2) might improve the accuracy of the genealogy.
Finally, future developments may account for geography in the HMM,
favouring attachment of samples to geographically close lineages.

>>>>

By automatically detecting and incorporating recombination events into the ARG,
sc2ts provides us with a systematic means of
discovering and assessing evidence around these events.
Table~\ref{tab:pango_x_lineages} summarizes
these results for the 44 Pango X lineages present in the ARG
(XAC and XAD have been omitted, and XM partially represented,
corresponding to 23 omitted samples in total;
see Document~\ref{sec:supplement_x_pango_lineages}).
For 16 Pango X lineages, the oldest node
that has been assigned to the focal Pango lineage is a recombination node, and
that node is the ancestor of all samples
assigned to that (or a descendant) Pango lineage.
We designate these as Type I events, where the ARG reflects the
hierarchical structure of Pango designation.
Among these events, concordance with the community-consensus 
parent lineages and breakpoint intervals is excellent,
with similar performance to
the RecombinHunt \cite{Alfonsi2024}
and CovRecomb \cite{Li2024CovRecomb} detection methods
(STAR Methods; Table~\ref{tab:method_concordance}; see also
Document~\ref{sec:jackson_recombs} for analysis of the 
Jackson et al.\ recombinants).
Of the Type I events, 13 are monophyletic for the focal Pango lineage
(i.e., the descendants are entirely from that lineage);
Figure~\ref{fig:subgraphs}A shows XA as an example.

@@

We validated the recombination events in the sc2ts ARG by running 
3SEQ~\cite{Lam2018improved} on the putative recombinant 
sequence trios (STAR Methods), 
and 484 of the 855 (57\%) were classified as recombinant.
Validation status correlates strongly with our QC filter, with 
95\% of the QC-passing recombinants 
and 29\% of the QC-failing recombinants
being classified as recombinant.
In addition to 3SEQ, we ran the CovRecomb~\cite{Li2024CovRecomb} 
and rebar recombination detection methods on the putative recombinant 
sequences, summarized in Table~\ref{tab:recombinant_validation}.
CovRecomb has a substantially lower validation rate at 25\% 
among the QC-passing recombinants. This is expected, as 
CovRecomb and is specifically intended to detect interlineage recombinants
and has previously been shown to be
less sensitive than other methods~\cite{Li2024CovRecomb}.
Rebar validation is intermediate, at 57\% among QC-passing recombinants.
(See also Figures~\ref{fig:recombinant_evidence_cov_recomb},
and \ref{fig:recombinant_evidence_rebar}.)
% We don't talk about RIPPLES results yet as we haven't discussed it, a
% and that would be confusing.

@@

Recombination is rare in SARS-CoV-2, and
ultimately judgments about plausibility require expert analysis.
To aid such interpretation and help identify quality control issues
we developed a visualization tool
inspired by SnpIt~\cite{OToole2024} and 
the recombinant view of RIVET~\cite{Smith2023tracking},
showing the recombination-informative sites
between a recombinant and its parent sequences, and
highlighting potentially problematic loci.
Each copying pattern shows the positions
where the allelic state in a recombinant
(middle row) matches that of the left (upper row, colored green) 
or right parent (lower row, colored blue),
or where the recombination event requires a \emph{de novo} mutation
(gold, with mutational change below). [should we also mention the red/orange lines under the patterns that identify adjacent sites?]
Figure~\ref{fig:recombinant_evidence}D shows some example copying patterns,
illustrating this spectrum of plausibility.
XA is an example of a highly plausible recombinant, shown by the 
large number of supporting loci on the left and right. As we have 
only one \emph{de novo} mutation, we can be sure that the parents are closely
observed [Some details about XA here, how well the parents were sampled etc].
XBB is also strongly supported by many loci on the left and right, 
but the presence of many \emph{de novo} mutations
observed in the copying patterns suggests the existence of
many genetically distinct, unsampled ancestors 
[FIXME: in this case, probably BJ.1, see issue 336],
and some caution should be exercised in interpreting them. 
See Figures~\ref{fig:recombinant_evidence}D and
\ref{fig:pango_x_copying_patterns}
for copying patterns of the Pango X lineages, and
Figure~\ref{fig:copying_pattern_examples} for copying patterns
highlighting specific types of QC problems
(and more details on the visualizations).

@@

In terms of \emph{detecting} recombination events, 
sc2ts and the combination of UShER and RIPPLES have much in common, 
and it is likely that very similar results could be obtained if run
in the same way. That is, if we constructed an UShER tree by 
incrementally adding the same daily batches as used for sc2ts
and examined any newly-added long branches each day
using RIPPLES, we would likely get very similar results to sc2ts
in terms of recombinants detected.
For a given matching operation, RIPPLES and the LS HMM 
(under its current parameterisation) 
are deterministically solving the same well-defined problem
(albeit with radically different implementations) and they 
must therefore produce similar answers. 
Where sc2ts and UShER+RIPPLES diverge, however, is in terms 
of how well such signals of recombination can be preserved
over longer timescales. Continuing our thought experiment
in which we run UShER and RIPPLES in daily batches, suppose
that on a given day a branch with 4 mutations is identified 
as being a recombinant. Unless this branch is specifically 
``protected'' as representing a recombinant it may 
be broken up by subsequent parsimony operations, thus 
erasing the signal of recombination. This would appear
to be the mostly likely explanation for the ``missing'' 
X-lineages like XBB (detected by sc2ts at $k=4$ but 
not by UShER+RIPPLES at $p=4$).

@@

% P2. This is a conservative lower-bound on recombination. 
% Discuss other estimates.
% 
% SHZ: Why should we care about getting this number/frequency right?
% Sure, popgen folks love estimating this, but to other people,
% existing estimates already tell us that
% recombination in SARS-CoV-2 is (qualitatively) rare,
% even without the advances of sc2ts.
% Also, we are talking about the frequency of detectable recombination
% in the last sentence, right?
% The true frequencey of recombination is unknowable from this type of data.
% So, the best we are ever going to get is a better lower bound.
The recombination frequency estimated from the \texttt{sc2ts} ARG is
substantially lower than that implied by the combination of UShER+RIPPLES,
largely reflecting differences in model formulation and parameter choice.
RIPPLES, applied post-hoc to UShER trees of 1.6 million genomes up to May 2021,
reported 589 unique recombination events with 43,104 descendant
samples---corresponding to roughly 2.7\% of all genomes being of recombinant
origin.  Using the same time interval, our inference at \(k=4\) identifies only
14 high-confidence recombination events with 172 descendants among 501,235
genomes (0.034\%), a 40- to 80-fold reduction.  This difference is largely
explained by the stricter parsimony penalty adopted here: increasing the
penalty from \(p=3\) (as used in the RIPPLES analysis) 
to \(k=4\) greatly reduces false
positives while retaining nearly all verified recombinant lineages, 
including those missed by RIPPLES at equivalent thresholds.  
We argue that the post-hoc nature of UShER+RIPPLES can erase evidence 
of recombination as
tree rearrangements proceed, thereby requiring the more lenient 
threshold in order to find known recombinats.
These comparisons suggest that our estimate represents a conservative 
lower bound. Further work will be required to determine
the true frequency of recombination in SARS-CoV-2 evolution.

% P3. Clarifies relationships between pango X lineages, simplifying
% the Omicron sublineage.
% 
% SHZ: I don't think knowing the relationships between Pango Xs is what's really important.
% Sure, phylogeneticists would get a kick out of it,
% but I don't think the majority of Cell readership would.
% I think what is important is knowing
% how many emerging recombinants carrying new combinations of mutations of concern
% we are actually dealing with in real time.
% Clarifying the relationships between Pango Xs make this possible.

@@

% P4. Results on global case numbers and frequency of recomb.
% SHZ: Should we mention how the reliability of global case number data
% and variation in sampling effort over time and across countries
% might affect our interpretation of these findings?
[FIXME: This is direct output from ChatGPT.  Review!]
The dated ARG allows us to place recombination within its epidemiological
context, revealing clear temporal structure in when recombinant lineages arise.
The frequency of inferred recombination events follows global case numbers
through the pandemic, with peaks during the Delta and Omicron waves when both
infection prevalence and genetic diversity were high.  Quantitatively, the best
single predictor of weekly recombinant counts is the product of global cases
and lineage heterozygosity (\(nH\)), which explains nearly a quarter of the
variation in event frequency (\(R^{2}_{\mathrm{adj}} = 0.24\)).  This
relationship is consistent with the mechanistic expectation that recombination
requires both frequent co-infections (driven by high case numbers) and
genetically distinct co-infecting viruses (driven by lineage diversity).  In
other words, recombination becomes most detectable when epidemiological and
evolutionary opportunity coincide.  These results imply that the emergence of
novel recombinants is not uniform through time but concentrated into
predictable ``hot periods'' during large, antigenically diverse outbreaks, when
the ecological conditions for viral recombination are most favourable.

% P5. Confirm breakpoints around Spike. 
% SHZ: Maybe we can say it's a known (instead of robust)
% pattern of detectable recombination in CoVs in general?
% There are some citations for template switching hotspots in SC2.
% Also, it's not only about selection favoring new combinations of mutations in Spike.
% It's about combinations of mutations in, say, ORF1ab, getting selected against.
% Maybe mention epistasis?
[FIXME: This is directish output from ChatGPT. Review!]
The genomic distribution of breakpoints in the ARG is highly uneven, with a
pronounced enrichment near the $5'$ boundary of the Spike gene and across the
S1–S2 junction.  This pattern closely mirrors that observed in the
UShER+RIPPLES dataset \citep{Turakhia2022} and in earlier 
analyses \citep{Jackson2021} [more cites], but here it is recovered
independently over a much longer sampling period and under a more conservative
inference regime.  Under our more conservative estimate of the overall
prevalence of recombination, we find that this enrichment persists despite a substantial
reduction in the total number of inferred recombination events, indicating that
it reflects a genuine and stable feature of SARS-CoV-2 evolution rather than a
by-product of detection sensitivity.  Across all high-confidence events, the
median breakpoint interval length is $\sim 2.2$kb and declines with parental
divergence, consistent with expectations from sequence-based localization
accuracy.  These results show that even with a conservative lower bound on
recombination frequency, the concentration of breakpoints around Spike remains
a robust signature of coronavirus genome biology—likely reflecting a
combination of mechanistic biases in template switching and selective
constraints on Spike function.

% P6. Sc2ts method is a major advance. We validate with real data
% rather than simulations. Forward looking statement about simulations.
% 
% SHZ: Could at least some of this go into the Introduction?
In the broader context of pathogen genomics, \texttt{sc2ts} demonstrates that
it is now possible to reconstruct recombination-aware genealogies at the scale
of entire pandemics.  Although recombination is recognised as a major
evolutionary force in many pathogens—including RNA viruses such as influenza
and HIV, and numerous bacterial species—genealogical analyses have typically
relied on tree-based approximations that ignore or collapse recombination.
This simplification has limited the accuracy of downstream evolutionary and
epidemiological inference, even in well-studied systems.  \texttt{Sc2ts}
overcomes this barrier by directly inferring an ancestral recombination graph
from millions of viral genomes, producing the first coherent reticulate
genealogy for any pathogen dataset of comparable size.  The ability to
integrate recombination into genome-wide genealogies at this scale represents a
major conceptual step forward: it transforms recombination from a confounding
complication into an analyzable component of pathogen evolution.

@@

We close by emphasizing several current limitations and potential directions to improve sc2ts.
First, more stringent filters to exclude false positive recombinants during inference
would obviate post-hoc treatment of these recombinants.
Second, indels carry important phylogenetic signals,
for example, as defining mutations of Alpha and Delta (REF).
We currently treat multi-base deletions as missing data or multiple independent deletions,
but they are in reality single events.
More sophisticated treatments of multi-base indels in the HMM could help recover additional signals
(and avoid false signals) of recombination.
Third, the current HMM assumes a simplistic mutation model,
which does not account for documented transition-to-transversion rate bias (REF)
and context-dependent variation in mutation rate (REF).
A more realistic mutation model may help improve inference of local genetic relationships in the ARG.
Fourth, an inherent limitation of detecting recombination from sequence data,
which applies to all methods, is that it is impossible to detect events with few (or no) supporting bases,
as these cannot be distinguished from mutation events.
This limitation makes it particularly challenging to detect events near the termini of linear viral genomes.
Fifth, it is challenging to accurately reconstruct global events
when there are regions with poor sampling and jurisdictions with long delays
before sequences become publicly available (REF).
Revisiting the ARG topology periodically
(e.g., rewinding the ARG a month and repeating with samples that became available later)
and better handling of putative recombinants with suspiciously long branches
(e.g., Delta and Omicron BA.2) might improve the accuracy of the genealogy.
Finally, future developments may account for geography in the HMM,
favoring attachment of samples to geographically close lineages.
